Synthesising Termination Proofs for Bit-Vector Programs

### 2014-07-07 papers/termination/paper.tex
nudges

We present a reduction from program \emph{termination} to program
\emph{synthesis} via a fragment of second-order logic.  This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination of fixed-width and floating-point
arithmetic programs.
%
\end{abstract}

@@

{\bf Limitations.}
Our algorithm proves termination for transition systems with finite state
spaces.  The (non-)termination proofs take the form of ranking functions and
program invariants that are expressed in a quantifier free language.  This
formalism is powerful enough to handle a large fragment of C, but is not
rich enough to analyse code which uses the heap or arrays.  Also, we have not
yet added support for recursion to our encoding.

@@

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano
arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
In practice, the undefined behaviour is implemented just as if the
arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its
machine-arithmetic result is the same as the result of the same operation
performed on integers.

The final source of difference between integer arithmetic and bitvector
arithmetic stems from width conversions.  Many programming languages allow
numeric variables of different types, which can be represented using words
of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
hold the value $0$ after the following code is executed:
%
\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} . \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.
%
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all quantified variables range over vectors.

@@

We will begin our discussion by showing how to encode in the synthesis fragment the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
G states $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Note that different formulations for unconditional termination are possible. 
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive
closure~\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes
that the second-order logic has a primitive predicate for disjunctive
well-foundedness.  In contrast, our formulation in Definition~\ref{def:UT}
does not use a primitive disjunctive well-foundedness predicate.  \\

@@

So if our transition relation is deterministic we can say, without
loss of generality, that non-termination of the loop is equivalent
to the existence of a closed recurrence set.  However if $T$ is
non-deterministic, it may be that there is an open recurrence
set but not closed recurrence set.  To see this, consider the following
loop:
%
\begin{lstlisting}[language=C]
while(x != 0) {
  y = nondet();
  x = x-y;
}
\end{lstlisting}

It is clear that this loop has many non-terminating executions,
e.g. the execution where \lstinline!nondet()! always returns 0.
However it is equally clear that each state has a successor
that exits the loop, i.e. when \lstinline|nondet()| returns
the value currently stored in \lstinline|x|.  So this loop
has an open recurrence set, but no closed recurrence set
and hence we cannot specify a proof of its non-termination
with formula {\bf [CNT]}.

@@

Termination for bitvector programs is known to be
PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
expected results for soundness and completeness of our algorithm for
deciding it.
%
\begin{theorem}\label{thm:synth-sound}
Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$, then
$P \models \sigma$.
\end{theorem}

>>>>

We present a reduction from program \emph{termination} to program
\emph{synthesis} via a fragment of second-order logic.  This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination analysis of programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

{\bf Limitations.} Our algorithm proves termination for transition systems
with finite state spaces.  The (non-)termination proofs take the form of
ranking functions and program invariants that are expressed in a
quantifier-free language.  This formalism is powerful enough to handle a
large fragment of C, but is not rich enough to analyse code that uses the
heap or unbounded arrays.  Also, we have not yet added support for recursion
to our encoding.

@@

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano
arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
In practice, the undefined behaviour is implemented just as if the
arithmetic had been unsigned.}  To give an example, $(2^k - 1) + 1 \equiv 0
\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its
machine-arithmetic result is the same as the result of the same operation
performed on integers.

The final source of difference between integer arithmetic and bit-vector
arithmetic stems from width conversions.  Many programming languages allow
numeric variables of different types, which can be represented using words
of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
hold the value $0$ after the following code is executed:
%
\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} .\, \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over functions, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.
%
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all quantified variables range over vectors.

@@

We will begin our discussion by showing how to encode in the synthesis fragment the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
States $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

Note that different formulations for unconditional termination are possible. 
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive
closure~\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes
that the second-order logic has a primitive predicate for disjunctive
well-foundedness.  By contrast, our formulation in Definition~\ref{def:UT}
does not use a primitive disjunctive well-foundedness predicate.  \\

@@

So if our transition relation is deterministic, we can say, without
loss of generality, that non-termination of the loop is equivalent
to the existence of a closed recurrence set.  However if $T$ is
non-deterministic, it may be that there is an open recurrence
set but not closed recurrence set.  To see this, consider the following
loop:
%
\begin{lstlisting}[language=C]
while(x != 0) {
  y = nondet();
  x = x-y;
}
\end{lstlisting}

It is clear that this loop has many non-terminating executions,
e.g. the execution where \lstinline!nondet()! always returns 0.
However it is equally clear that each state has a successor
that exits the loop, i.e. when \lstinline|nondet()| returns
the value currently stored in \lstinline|x|.  So this loop
has an open recurrence set, but no closed recurrence set
and hence we cannot give a proof of its non-termination
with formula {\bf [CNT]}.

@@

Termination for bit-vector programs is known to be
PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
expected results for soundness and completeness of our algorithm for
deciding it.
%
\begin{theorem}\label{thm:synth-sound}
Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$, then
$P \models \sigma$.
\end{theorem}
